---
title: "Main_Analysis"
---

# Summary

* Process the survey data for Bike Count. 
* Input: Single Excel file exported from Filemaker Pro database (987 rows and 47 columns). It is assumed that ONLY the rows for one single year are present
* Developed specifically for Bike Count 2019, but should work with any future Bike Count edition, possibly with a few minor adjustments.
* Output: Produces an Excel file where the data has been restructured and aggregated (XXX columns and XXX rows). 

* Last modified: 12/03/2019

# ** 1. Init

## Initialize the paths and names. 

```{r init}
library(readxl) 
library(xlsx)

home <- "C:/Users/delph/Desktop/GIS/BCGP/Bike_count/BikeCount2019/"
inFolder <- "Raw/"
inTable <- "test5"
outFolder <- "Output/"

```


# ** 2. Read data and prepare it

## Read the raw Excel input spreadsheet 

```{r read_excel}

# Create the path
fileToRead <- paste(home, inFolder, inTable, ".xlsx", sep="")
fileToRead

# Read the file
bcDataOrig <- as.data.frame(read_excel(fileToRead))

# Create a copy of the dataframe to keep the original intact
bcData <- bcDataOrig
bcData

```


## Do some cleanup and remove superfluous columns

```{r cleanup}

# Clean up the names of the columns
names(bcData) = gsub(pattern = "IntersectionsPlusDirections::", 
                     replacement = "", x = names(bcData))
names(bcData) = gsub(pattern = "Derived ", 
                     replacement = "", x = names(bcData))
names(bcData) = gsub(pattern = "Count (.+)$", 
                     replacement = "\\1 Count", x = names(bcData))
names(bcData)[names(bcData)=="__CountOccurrencePK"] <- "Record ID"
names(bcData)[names(bcData)=="_IntersectionsFK"] <- "Inters And Direc ID"

# Add this new variable that we will populate later
bcData$"BPH per Intersection" = 0

# Reorder the columns and drop the ones we don't need
bcData2 <- bcData[c("Record ID",	"Inters And Direc ID", "Bridge vs Roads",	"Intersection",	"Street",	"Direction", "Date",	"Time",	"AM PM",	"Year",	"Facility for Street", "Total Count", "Bikes Per Hour Projected", "Num of Directions in Intersection",	"BPH per Intersection", "Male Count", "Female Count",  "With Traffic Male",	"With Traffic Female",	"Sidewalk Male", "Sidewalk Female", "Wrong Way Male",	"Wrong Way Female", 	"Helmet Male",	"Helmet Female",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses", "Pedestrians", "EScooters")]

# Replace NAs with 0
numOfCol <- ncol(bcData2)
bcData2[12:numOfCol][is.na(bcData2[12:numOfCol])] <- 0

bcData2

```

## Add intermediary columns

```{r add_intermediary_columns}

# Compute the Bicycle Per Hour equivalent per intersection for each row.
# This will be used to compute the total average per hour at every intersection
bcData2$"BPH per Intersection" <- bcData2$"Total Count" * 
  bcData2$"Num of Directions in Intersection" * 4

# Add columns for totals across genders
bcData2$"With Traffic" = bcData2$"With Traffic Male" +	bcData2$"With Traffic Female"
bcData2$"Sidewalk" = bcData2$"Sidewalk Male" +	bcData2$"Sidewalk Female"
bcData2$"With Traffic" = bcData2$"With Traffic Male" +	bcData2$"With Traffic Female"
bcData2$"Wrong Way" = bcData2$"Wrong Way Male" +	bcData2$"Wrong Way Female"
bcData2$"Helmet" = bcData2$"Helmet Male" +	bcData2$"Helmet Female"

bcData2
```


# ** 3. Compute Bikes Per Hour (BPH) numbers

## Make lists of names

```{r make_lists_of_names}

# Get the list of all bridge names
bridgeNames <- unique(bcData2[bcData2$"Bridge vs Roads" == "Bridge" ,]
                      $Intersection)

# Get the list of all street intersection names
streetIntersectionNames <- unique(bcData2[bcData2$"Bridge vs Roads" == "Road" ,]$Intersection)

# Get the list of all bicycle facility names
facilityNames <- unique(bcData2$"Facility for Street")

bridgeNames
streetIntersectionNames
facilityNames
```
## Compute BPH by bicycle facilities

```{r compute_BPH_facilities}

BPHSummary <- as.data.frame(matrix(ncol=0, nrow=1))
rownames(BPHSummary) <- c("BPH")

# For each facility type
for ( f in facilityNames){ 
  #Get the subset all all entry rows for that facility type:
  facilitySubset <- bcData2[bcData2$"Facility for Street" == f, ]

  BPHSummary[, f] <- round(sum(facilitySubset$"Bikes Per Hour Projected")/nrow(facilitySubset), 2)
}

BPHSummary

```


## Compute BPH for bridges

```{r compute_BPH_bridges}

# For each bridge 
for ( b in bridgeNames){ 
  #Get the subset of all entry rows for that bridge:
  bridgeSubset <- bcData2[bcData2$"Intersection" == b, ]
 
  BPHSummary[, b] <- round(sum(bridgeSubset$"BPH per Intersection")/nrow(bridgeSubset), 2)
}

BPHSummary
```


## Compute BPH for street intersections 

```{r compute_BPH_street_intersections}

# For each street intersection 
for ( i in streetIntersectionNames){ 
  #Get the subset of all entry rows for that intersection:
  intersectionSubset <- bcData2[bcData2$"Intersection" == i, ]
 
  BPHSummary[, i] <- round(sum(intersectionSubset$"BPH per Intersection")/nrow(intersectionSubset), 2)
}

BPHSummary

```

# ** 4. Compute Bikes Per Hour (BPH) numbers for individual streets

We put this in a different dataframe since it is different enough.

## Prepare subset and make lists of names

```{r make_lists_of_names}

# Create road subset
roadSubset <- bcData2[bcData2$"Bridge vs Roads" == "Road" ,]

# Add composed names for individual street names in the road subset, such as "Pine@21st"
roadSubset$PrettyStreetName <- roadSubset$Intersection
roadSubset$PrettyStreetName <- mapply(gsub, pattern=roadSubset$Street, 
                                   x=roadSubset$PrettyStreetName, replacement="")
roadSubset$PrettyStreetName <- gsub(pattern = "\\sand\\s", 
                     replacement = "", roadSubset$PrettyStreetName)
roadSubset$PrettyStreetName <- paste(roadSubset$Street, "(@", roadSubset$PrettyStreetName,")", sep="" )


# Get the list of all individual street names
individualStreetNames <- unique(roadSubset$PrettyStreetName)

roadSubset
individualStreetNames

```


## Compute BPH by street

```{r compute_BPH_by_street}

streetBPHSummary <- as.data.frame(matrix(ncol=0, nrow=1))
rownames(streetBPHSummary) <- c("individual street BPH")

# For each street  
for ( i in individualStreetNames){ 
  #Get the subset of all entry rows for that intersection:
  streetSubset <- roadSubset[roadSubset$PrettyStreetName == i, ]
 
  streetBPHSummary[, i] <- round(sum(
    streetSubset$"Bikes Per Hour Projected")/nrow(streetSubset), 2)
}

streetBPHSummary

```


# ** 5. Bicyclist behavior

## Compute aggregated totals by behavior

```{r compute_totals_by_behavior}

behaviorSummary <- as.data.frame(matrix(ncol=0, nrow=20))

rownames(behaviorSummary) <- c("Total Count", "Male Count", "Female Count",  "With Traffic", "With Traffic Male",	"With Traffic Female",	"Sidewalk", "Sidewalk Male", "Sidewalk Female", "Wrong Way", "Wrong Way Male",	"Wrong Way Female", 	"Helmet", "Helmet Male",	"Helmet Female",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses", "Pedestrians", "EScooters")

behaviorSummary$Totals <- colSums(bcData2[c("Total Count", "Male Count", "Female Count",  "With Traffic", "With Traffic Male",	"With Traffic Female",	"Sidewalk", "Sidewalk Male", "Sidewalk Female", "Wrong Way", "Wrong Way Male",	"Wrong Way Female", 	"Helmet", "Helmet Male",	"Helmet Female",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses", "Pedestrians", "EScooters")])

grandTotal <- behaviorSummary["Total Count", "Totals"]
grandTotal

behaviorSummary$Percent <- round((behaviorSummary$Totals / grandTotal)*100, digits = 2)
behaviorSummary

```

## Compute behaviors by bicycle facilities

```{r compute_behaviors_by_facilities}

# For each facility type
for ( f in facilityNames){ 
  #Get the subset all all entry rows for that facility type:
  facilitySubset <- bcData2[bcData2$"Facility for Street" == f, ]
  
  # Turn the Record ID column into the row names
  rownames(facilitySubset) <- facilitySubset$"Record ID"
  
  # Compute the column totals for the subsets and add them to the 
  # behaviorSummary dataset in a new column with the name of the bike facility (f)
  behaviorSummary[, f] <- colSums(facilitySubset[c("Total Count",  
                          "Male Count", 
                          "Female Count",  "With Traffic", "With Traffic Male",	
                          "With Traffic Female",	"Sidewalk", "Sidewalk Male", 
                          "Sidewalk Female", "Wrong Way", "Wrong Way Male",	
                          "Wrong Way Female", 	"Helmet", "Helmet Male",	
                          "Helmet Female",	"Bikes Share Users", 	
                          "Bike Share Sidewalk Wrong Way",	"Bikes on Buses",  "Pedestrians", "EScooters")])

  # Retrieve the total bike count for the current bike facility
  subGrandTotal <- behaviorSummary[1, f]

  # Creare a new column called "[bike facility name] Percent", and put in it the %s
  fieldName <- paste(f, "Percent", sep=" ")
  behaviorSummary[, fieldName] <- round((behaviorSummary[, f] / subGrandTotal)*100, digits = 2)

  }

#myres <- behaviorSummary[behaviorSummary$VarName =="Sidewalk",]
#myres <- behaviorSummary[behaviorSummary$VarName =="Helmet",]
#myres <- behaviorSummary[behaviorSummary$VarName =="Wrong Way",]

```


## Generate short behavior summary

```{r compute_subtotals}

shortBehaviorSummary <- behaviorSummary

shortBehaviorSummary <- shortBehaviorSummary[c("Total Count", "Male Count", "Female Count",  "With Traffic", "Sidewalk", "Wrong Way", "Helmet",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses", "Pedestrians", "EScooters"),]

shortBehaviorSummary

```

# ** 6. Create Excel output

## Initialize Excel workbook and set it up 

```{r export_to_excel_start}

# Put the file path together
fileToWrite <- paste(home, outFolder, inTable, "_processed.xlsx", sep="")
fileToWrite

# Create the workbook
wb = createWorkbook()

# Define styles for the data table row/column names
TABLE_ROWNAMES_STYLE <- CellStyle(wb) + Font(wb, isBold=TRUE)
TABLE_COLNAMES_STYLE <- CellStyle(wb) + Font(wb, isBold=TRUE) +
  Alignment(wrapText=TRUE, horizontal="ALIGN_CENTER") 

# Define some sheets ahead of time to make sure they appear first in the workbook
shortBehaviorSummarySheet = createSheet(wb, "Behavior Aggregation")
BPHSummarySheet = createSheet(wb, "BPH (Facilities & Intersec.)")
streetBPHSummarySheet = createSheet(wb, "BPH (Streets)")

```


## Add bicycle facility subset to the Excel workbook 

```{r export_facilities}

# For each facility type
for ( f in facilityNames){ 
  #Get the subset all all entry rows for that facility type:
  facilitySubset <- bcData2[bcData2$"Facility for Street" == f, ]
  
  # Put that subset into the Excel workbook as a new sheet for export purposes
  facilitySubsetSheet <- createSheet(wb, f)
  
  addDataFrame(facilitySubset, sheet=facilitySubsetSheet, startColumn=1, 
               row.names=FALSE, colnamesStyle = TABLE_COLNAMES_STYLE, 
               rownamesStyle = TABLE_ROWNAMES_STYLE)
}
```


## Add bridge subset to the Excel workbook 

```{r export_bridges}

# For each bridge 
 for ( b in bridgeNames){ 
  #Get the subset of all entry rows for that bridge:
  bridgeSubset <- bcData2[bcData2$"Intersection" == b, ]
 
  # Put that subset into the Excel workbook as a new sheet 
  bridgeSheet <- createSheet(wb, b)
  addDataFrame(bridgeSubset, sheet=bridgeSheet, startColumn=1, 
               row.names=FALSE, colnamesStyle = TABLE_COLNAMES_STYLE, 
               rownamesStyle = TABLE_ROWNAMES_STYLE)
}

```


## Add street intersection subset to the Excel workbook 

```{r export_street_intersection}

# For each street intersection 
for ( i in streetIntersectionNames){ 
  #Get the subset of  all entry rows for that street intersection:
  streetIntersectionSubset <- bcData2[bcData2$"Intersection" == i, ]
  
  # Put that subset into the Excel workbook as a new sheet for export purposes
  streetIntersectionSheet <- createSheet(wb, i)
  addDataFrame(streetIntersectionSubset, sheet=streetIntersectionSheet, 
               startColumn=1, row.names=FALSE, colnamesStyle = 
               TABLE_COLNAMES_STYLE, rownamesStyle = TABLE_ROWNAMES_STYLE)

}
```



## Export to Excel - End

```{r export_to_excel_end}

addDataFrame(shortBehaviorSummary, sheet=shortBehaviorSummarySheet, startColumn=1, row.names=TRUE, colnamesStyle = TABLE_COLNAMES_STYLE, rownamesStyle = TABLE_ROWNAMES_STYLE)

addDataFrame(BPHSummary, sheet=BPHSummarySheet, startColumn=1, row.names=TRUE, colnamesStyle = TABLE_COLNAMES_STYLE, rownamesStyle = TABLE_ROWNAMES_STYLE)

addDataFrame(streetBPHSummary, sheet=streetBPHSummarySheet, startColumn=1,row.names=TRUE, colnamesStyle = TABLE_COLNAMES_STYLE, rownamesStyle = TABLE_ROWNAMES_STYLE, )
#setColWidths(wb, streetBPHSummary, 1:ncol(streetBPHSummary), widths = 20)

bcDataSheet = createSheet(wb, "All Data - Cleaned")
addDataFrame(bcData2, sheet=bcDataSheet, startColumn=1, row.names=FALSE, colnamesStyle = TABLE_COLNAMES_STYLE, rownamesStyle = TABLE_ROWNAMES_STYLE)


bcDataOrigSheet = createSheet(wb, "Original Export")
addDataFrame(bcDataOrig, sheet=bcDataOrigSheet, startColumn=1, row.names=FALSE, colnamesStyle = TABLE_COLNAMES_STYLE, rownamesStyle = TABLE_ROWNAMES_STYLE)


saveWorkbook(wb, fileToWrite)

```













