---
title: "Main_Analysis"
---

# Summary

* Process the survey data for Bike Count. 
* Input: Single Excel file exported from Filemaker Pro database (987 rows and 47 columns). It is assumed that ONLY the rows for one single year are present
* Developed specifically for Bike Count 2019, but should work with any future Bike Count edition, possibly with a few minor adjustments.
* Output: Produces an Excel file where the data has been restructured and aggregated (XXX columns and XXX rows). 

* Last modified: 11/30/2019

# Init

Initialize the paths and names. 

```{r init}
library(readxl) 
library(xlsx)

home <- "C:/Users/delph/Desktop/GIS/BCGP/Bike_count/BikeCount2019/"
inFolder <- "Raw/"
inTable <- "test5"
outFolder <- "Output/"

```

## Initialize Excel export output 

We will build the content little by little throughout the script.

```{r export_to_excel_start}

fileToWrite <- paste(home, outFolder, inTable, "_3_processed.xlsx", sep="")
fileToWrite

wb = createWorkbook()

# Styles for the data table row/column names
TABLE_ROWNAMES_STYLE <- CellStyle(wb) + Font(wb, isBold=TRUE)
TABLE_COLNAMES_STYLE <- CellStyle(wb) + Font(wb, isBold=TRUE) +
  Alignment(wrapText=TRUE, horizontal="ALIGN_CENTER") 

# Defining some sheets ahead of time to make sure they appear first in the workbook
shortSummarySheet = createSheet(wb, "Short Summary")
fullSummarySheet = createSheet(wb, "Full Summary")

```


## Read the raw Excel table 

```{r read_excel}

# Create the path
fileToRead <- paste(home, inFolder, inTable, ".xlsx", sep="")
fileToRead

# Read the file
bcDataOrig <- as.data.frame(read_excel(fileToRead))

# Create a copy of the dataframe to keep the original intact
bcData <- bcDataOrig
bcData

```


## Do some cleanup and remove superfluous columns

```{r cleanup}

# Clean up the names of the columns
names(bcData) = gsub(pattern = "IntersectionsPlusDirections::", 
                     replacement = "", x = names(bcData))
names(bcData) = gsub(pattern = "Derived ", 
                     replacement = "", x = names(bcData))
names(bcData) = gsub(pattern = "Count (.+)$", 
                     replacement = "\\1 Count", x = names(bcData))
names(bcData)[names(bcData)=="__CountOccurrencePK"] <- "Record ID"
names(bcData)[names(bcData)=="_IntersectionsFK"] <- "Inters And Direc ID"

# Add this new variable that we will populate later
bcData$"BPH per Intersection" = 0

# Reorder the columns and drop the ones we don't need
bcData2 <- bcData[c("Record ID",	"Inters And Direc ID", "Bridge vs Roads",	"Intersection",	"Street",	"Direction", "Date",	"Time",	"AM PM",	"Year",	"Facility for Street", "Total Count", "Bikes Per Hour Projected", "Num of Directions in Intersection",	"BPH per Intersection", "Male Count", "Female Count",  "With Traffic Male",	"With Traffic Female",	"Sidewalk Male", "Sidewalk Female", "Wrong Way Male",	"Wrong Way Female", 	"Helmet Male",	"Helmet Female",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses", "Pedestrians", "EScooters")]

# Replace NAs with 0
numOfCol <- ncol(bcData2)
bcData2[12:numOfCol][is.na(bcData2[12:numOfCol])] <- 0

```

# Add intermediary columns

```{r add_intermediary_columns}

# Compute the Bicycle Per Hour equivalent per intersection for each row.
# This will be used to compute the total average per hour at every intersection
bcData2$"BPH per Intersection" <- bcData2$"Total Count" * bcData2$"Num of Directions in Intersection" * 4

# Add columns for totals across genders
bcData2$"With Traffic" = bcData2$"With Traffic Male" +	bcData2$"With Traffic Female"
bcData2$"Sidewalk" = bcData2$"Sidewalk Male" +	bcData2$"Sidewalk Female"
bcData2$"With Traffic" = bcData2$"With Traffic Male" +	bcData2$"With Traffic Female"
bcData2$"Wrong Way" = bcData2$"Wrong Way Male" +	bcData2$"Wrong Way Female"
bcData2$"Helmet" = bcData2$"Helmet Male" +	bcData2$"Helmet Female"

bcData2
```

## Compute totals

```{r compute_totals}


fullSummary <- as.data.frame(matrix(ncol=0, nrow=18))

rownames(fullSummary) <- c("Total Count", "Male Count", "Female Count",  "With Traffic", "With Traffic Male",	"With Traffic Female",	"Sidewalk", "Sidewalk Male", "Sidewalk Female", "Wrong Way", "Wrong Way Male",	"Wrong Way Female", 	"Helmet", "Helmet Male",	"Helmet Female",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses")

fullSummary$Totals <- colSums(bcData2[c("Total Count", "Male Count", "Female Count",  "With Traffic", "With Traffic Male",	"With Traffic Female",	"Sidewalk", "Sidewalk Male", "Sidewalk Female", "Wrong Way", "Wrong Way Male",	"Wrong Way Female", 	"Helmet", "Helmet Male",	"Helmet Female",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses")])

grandTotal <- fullSummary["Total Count", "Totals"]
grandTotal

fullSummary$Percent <- round((fullSummary$Totals / grandTotal)*100, digits = 2)
fullSummary

```

## Compute Bikes Per Hour numbers (BPH)

```{r compute_BPH}

BPHSummary <- as.data.frame(matrix(ncol=0, nrow=1))
rownames(BPHSummary) <- c("BPH")

# Get the list of all bicycle facility types
facilities <- unique(bcData2$"Facility for Street")

# For each facility type
for ( f in facilities){ 
  #Get the subset all all entry rows for that facility type:
  facilitySubset <- bcData2[bcData2$"Facility for Street" == f, ]

  BPHSummary[, f] <- round(sum(facilitySubset$"Bikes Per Hour Projected")/nrow(facilitySubset), 2)
}


```

## Compute subtotals by bicycle facility type

```{r compute_subtotals_facilities}

# For each facility type
for ( f in facilities){ 
  #Get the subset all all entry rows for that facility type:
  facilitySubset <- bcData2[bcData2$"Facility for Street" == f, ]
  
  # Put that subset into the Excel workbook as a new sheet for export purposes
  facilitySubsetSheet <- createSheet(wb, f)
  addDataFrame(facilitySubset, sheet=facilitySubsetSheet, startColumn=1, 
               row.names=FALSE, colnamesStyle = TABLE_COLNAMES_STYLE, 
               rownamesStyle = TABLE_ROWNAMES_STYLE)

  # Turn the Record ID column into the row names
  rownames(facilitySubset) <- facilitySubset$"Record ID"
  
  # Compute the column totals for the subsets and add them to the 
  # fullSummary dataset in a new column with the name of the bike facility (f)
  fullSummary[, f] <- colSums(facilitySubset[c("Total Count",  
                          "Male Count", 
                          "Female Count",  "With Traffic", "With Traffic Male",	
                          "With Traffic Female",	"Sidewalk", "Sidewalk Male", 
                          "Sidewalk Female", "Wrong Way", "Wrong Way Male",	
                          "Wrong Way Female", 	"Helmet", "Helmet Male",	
                          "Helmet Female",	"Bikes Share Users", 	
                          "Bike Share Sidewalk Wrong Way",	"Bikes on Buses")])

  # Retrieve the total bike count for the current bike facility
  subGrandTotal <- fullSummary[1, f]

  # Creare a new column called "[bike facility name] Percent", and put in it the %s
  fieldName <- paste(f, "Percent", sep=" ")
  fullSummary[, fieldName] <- round((fullSummary[, f] / subGrandTotal)*100, digits = 2)

  }

#myres <- fullSummary[fullSummary$VarName =="Sidewalk",]
#myres <- fullSummary[fullSummary$VarName =="Helmet",]
#myres <- fullSummary[fullSummary$VarName =="Wrong Way",]

```


## Compute Bikes Per Hour numbers (BPH)

```{r compute_BPH_intersection}

# Get the list of all bicycle facility types
intersections <- unique(bcData2$"Intersection")
#intersections

# For each intersection 
for ( i in intersections){ 
  #Get the subset all all entry rows for that intersection:
  intersectionSubset <- bcData2[bcData2$"Intersection" == i, ]
 
  BPHSummary[, i] <- round(sum(intersectionSubset$"BPH per Intersection")/nrow(intersectionSubset), 2)
}


```


## Compute subtotals by intersection

```{r compute_subtotals_intersection}


# For each intersection 
for ( i in intersections){ 
  #Get the subset all all entry rows for that intersection:
  intersectionSubset <- bcData2[bcData2$"Intersection" == i, ]
  
  # Put that subset into the Excel workbook as a new sheet for export purposes
  intersectionSubsetSheet <- createSheet(wb, i)
  addDataFrame(intersectionSubset, sheet=intersectionSubsetSheet, startColumn=1, 
               row.names=FALSE, colnamesStyle = TABLE_COLNAMES_STYLE, 
               rownamesStyle = TABLE_ROWNAMES_STYLE)

  # Turn the Record ID column into the row names
  rownames(intersectionSubset) <- intersectionSubset$"Record ID"
  
  # Compute the column totals for the subsets and add them to the 
  # fullSummary dataset in a new column with the name of the intersection (f)
  fullSummary[, i] <- colSums(intersectionSubset[c("Total Count", "Male Count", 
                          "Female Count",  "With Traffic", "With Traffic Male",	
                          "With Traffic Female",	"Sidewalk", "Sidewalk Male", 
                          "Sidewalk Female", "Wrong Way", "Wrong Way Male",	
                          "Wrong Way Female", 	"Helmet", "Helmet Male",	
                          "Helmet Female",	"Bikes Share Users", 	
                          "Bike Share Sidewalk Wrong Way",	"Bikes on Buses")])

  # Retrieve the total bike count for the current intersection
  subGrandTotal <- fullSummary[1, i]

  # Creare a new column called "[intersection name] Percent", and put in it the %s
  fieldName <- paste(i, "Percent", sep=" ")
  fullSummary[, fieldName] <- round((fullSummary[, i] / subGrandTotal)*100, digits = 2)

  }
```


## Generate short summary

```{r compute_subtotals}

shortSummary <- fullSummary

shortSummary <- shortSummary[c("Total Count", "Male Count", "Female Count",  "With Traffic", "Sidewalk", "Wrong Way", "Helmet",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses"),]

# NOT USED
# Transpose rows into columns 
#shortSummary <- data.frame(t(shortSummary))

# Remove the periods that were added in the column names by default
#names(shortSummary) = gsub(pattern = "\\.", 
#                     replacement = " ", x = names(shortSummary))


shortSummary

```


## Export to Excel - End

```{r export_to_excel_end}

addDataFrame(shortSummary, sheet=shortSummarySheet, startColumn=1, row.names=TRUE, colnamesStyle = TABLE_COLNAMES_STYLE, rownamesStyle = TABLE_ROWNAMES_STYLE)

addDataFrame(fullSummary, sheet=fullSummarySheet, startColumn=1, row.names=TRUE, colnamesStyle = TABLE_COLNAMES_STYLE, rownamesStyle = TABLE_ROWNAMES_STYLE)

bcDataSheet = createSheet(wb, "All Data - Cleaned")
addDataFrame(bcData, sheet=bcDataSheet, startColumn=1, row.names=FALSE, colnamesStyle = TABLE_COLNAMES_STYLE, rownamesStyle = TABLE_ROWNAMES_STYLE)


bcDataOrigSheet = createSheet(wb, "Original Data")
addDataFrame(bcDataOrig, sheet=bcDataOrigSheet, startColumn=1, row.names=FALSE, colnamesStyle = TABLE_COLNAMES_STYLE, rownamesStyle = TABLE_ROWNAMES_STYLE)


saveWorkbook(wb, fileToWrite)

```













