---
title: "Main_Analysis"
---

# Summary

* Process the survey data for Bike Count. 
* Input: Single Excel file exported from Filemaker Pro database (987 rows and 47 columns). 
* Developed specifically for Bike Count 2019, but should work with any future Bike Count edition, possibly with a few minor adjustments.
* Output: Produces an Excel file where the data has been restructured and aggregated (XXX columns and XXX rows). 

* Last modified: 11/30/2019

# Init

Initialize the paths and names. 

```{r init}
library(readxl) 
#library(writexl) 

home <- "C:/Users/delph/Desktop/GIS/BCGP/Bike_count/BikeCount2019/"
inFolder <- "Raw/"
inTable <- "test4.xlsx"
outFolder <- "Output/"
```

## Read the raw Excel table 

```{r read_excel}
# Read the file
fileToRead <- paste(home, inFolder, inTable, sep="")
fileToRead
bcData <-read_excel(fileToRead)

bcData

```

## Do some cleanup and remove superfluous columns

```{r cleanup}

names(bcData) = gsub(pattern = "IntersectionsPlusDirections::", 
                     replacement = "", x = names(bcData))
names(bcData) = gsub(pattern = "Derived ", 
                     replacement = "", x = names(bcData))
names(bcData) = gsub(pattern = "Count (.+)$", 
                     replacement = "\\1 Count", x = names(bcData))
names(bcData)[names(bcData)=="__CountOccurrencePK"] <- "Record ID"
names(bcData)[names(bcData)=="_IntersectionsFK"] <- "Inters And Direc ID"

bcData$Intersection = ""
bcData$"Bike Count for Intersec (* 8 or * 4)" = 0

bcData2 <- bcData[c("Record ID",	"Inters And Direc ID", "Bridge vs Roads",	"Intersection",	"Street",	"Direction", "Date",	"Time",	"AM PM",	"Year",	"Facility for Street", "Total Count", "Bikes Per Hour Projected", 	"Bike Count for Intersec (* 8 or * 4)", "Male Count", "Female Count",  "With Traffic Male",	"With Traffic Female",	"Sidewalk Male", "Sidewalk Female", "Wrong Way Male",	"Wrong Way Female", 	"Helmet Male",	"Helmet Female",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses")]

names(bcData)

# Replace NAs with 0
bcData2[12:27][is.na(bcData2[12:27])] <- 0


```


## Compute totals

```{r compute_totals}

# Add columns for totals across genders
bcData2$"With Traffic" = bcData2$"With Traffic Male" +	bcData2$"With Traffic Female"
bcData2$"Sidewalk" = bcData2$"Sidewalk Male" +	bcData2$"Sidewalk Female"
bcData2$"With Traffic" = bcData2$"With Traffic Male" +	bcData2$"With Traffic Female"
bcData2$"Wrong Way" = bcData2$"Wrong Way Male" +	bcData2$"Wrong Way Female"
bcData2$"Helmet" = bcData2$"Helmet Male" +	bcData2$"Helmet Female"

fullSummary <- as.data.frame(matrix(ncol=1, nrow=18))
fullSummary$VarName <- c("Total Count", "Male Count", "Female Count",  "With Traffic", "With Traffic Male",	"With Traffic Female",	"Sidewalk", "Sidewalk Male", "Sidewalk Female", "Wrong Way", "Wrong Way Male",	"Wrong Way Female", 	"Helmet", "Helmet Male",	"Helmet Female",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses")
fullSummary$Totals <- colSums(bcData2[c("Total Count", "Male Count", "Female Count",  "With Traffic", "With Traffic Male",	"With Traffic Female",	"Sidewalk", "Sidewalk Male", "Sidewalk Female", "Wrong Way", "Wrong Way Male",	"Wrong Way Female", 	"Helmet", "Helmet Male",	"Helmet Female",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses")])

  grandTotal <- fullSummary[1, "Totals"]
grandTotal

fullSummary$Percent <- fullSummary$Totals / grandTotal

fullSummary

#Add BPH based on Bikes Per Hour Projected???

```

## Compute subtotals

```{r compute_subtotals}

facilities <- unique(bcData2$"Facility for Street")

for ( f in facilities){ 
  facilitySubset <- bcData2[bcData2$"Facility for Street" == f, ]
  
  fullSummary[, f] <- colSums(facilitySubset[c("Total Count", "Male Count", "Female Count",  "With Traffic", "With Traffic Male",	"With Traffic Female",	"Sidewalk", "Sidewalk Male", "Sidewalk Female", "Wrong Way", "Wrong Way Male",	"Wrong Way Female", 	"Helmet", "Helmet Male",	"Helmet Female",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses")])

  subGrandTotal <- fullSummary[1, f]

  fieldName <- paste(f, "Percent", sep=" ")
  
  fullSummary[, fieldName] <- fullSummary[, f] / subGrandTotal

  #Add BPH based on Bikes Per Hour Projected???

  }

myres <- fullSummary[fullSummary$VarName =="Sidewalk",]
myres <- fullSummary[fullSummary$VarName =="Helmet",]
myres <- fullSummary[fullSummary$VarName =="Wrong Way",]


```


