---
title: "Main_Analysis"
---

# Summary

* Process the survey data for Bike Count. 
* Input: Single Excel file exported from Filemaker Pro database (987 rows and 47 columns). 
* Developed specifically for Bike Count 2019, but should work with any future Bike Count edition, possibly with a few minor adjustments.
* Output: Produces an Excel file where the data has been restructured and aggregated (XXX columns and XXX rows). 

* Last modified: 11/30/2019

# Init

Initialize the paths and names. 

```{r init}
library(readxl) 
library(xlsx)

home <- "C:/Users/delph/Desktop/GIS/BCGP/Bike_count/BikeCount2019/"
inFolder <- "Raw/"
inTable <- "test4"
outFolder <- "Output/"
```

## Export output to Excel - Start

```{r export_to_excel_start}

fileToWrite <- paste(home, outFolder, inTable, "_2_processed.xlsx", sep="")
fileToWrite

wb = createWorkbook()

# Styles for the data table row/column names
TABLE_ROWNAMES_STYLE <- CellStyle(wb) + Font(wb, isBold=TRUE)
TABLE_COLNAMES_STYLE <- CellStyle(wb) + Font(wb, isBold=TRUE) +
  Alignment(wrapText=TRUE, horizontal="ALIGN_CENTER") 

# Defining some sheets ahead of time to make sure they appear first in the workbook
shortSummarySheet = createSheet(wb, "Short Summary")
fullSummarySheet = createSheet(wb, "Full Summary")

```


## Read the raw Excel table 

```{r read_excel}
# Read the file
fileToRead <- paste(home, inFolder, inTable, ".xlsx", sep="")
fileToRead

bcDataOrig <- as.data.frame(read_excel(fileToRead))
bcData <- bcDataOrig

```

## Do some cleanup and remove superfluous columns

```{r cleanup}

names(bcData) = gsub(pattern = "IntersectionsPlusDirections::", 
                     replacement = "", x = names(bcData))
names(bcData) = gsub(pattern = "Derived ", 
                     replacement = "", x = names(bcData))
names(bcData) = gsub(pattern = "Count (.+)$", 
                     replacement = "\\1 Count", x = names(bcData))
names(bcData)[names(bcData)=="__CountOccurrencePK"] <- "Record ID"
names(bcData)[names(bcData)=="_IntersectionsFK"] <- "Inters And Direc ID"

#Multiplier For Intersection

bcData$Intersection = ""
bcData$"Bike Count for Intersec (* 8 or * 4)" = 0

bcData2 <- bcData[c("Record ID",	"Inters And Direc ID", "Bridge vs Roads",	"Intersection",	"Street",	"Direction", "Date",	"Time",	"AM PM",	"Year",	"Facility for Street", "Total Count", "Bikes Per Hour Projected", 	"Bike Count for Intersec (* 8 or * 4)", "Male Count", "Female Count",  "With Traffic Male",	"With Traffic Female",	"Sidewalk Male", "Sidewalk Female", "Wrong Way Male",	"Wrong Way Female", 	"Helmet Male",	"Helmet Female",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses")]

names(bcData)

# Replace NAs with 0
bcData2[12:27][is.na(bcData2[12:27])] <- 0


```


## Compute totals

```{r compute_totals}

# Add columns for totals across genders
bcData2$"With Traffic" = bcData2$"With Traffic Male" +	bcData2$"With Traffic Female"
bcData2$"Sidewalk" = bcData2$"Sidewalk Male" +	bcData2$"Sidewalk Female"
bcData2$"With Traffic" = bcData2$"With Traffic Male" +	bcData2$"With Traffic Female"
bcData2$"Wrong Way" = bcData2$"Wrong Way Male" +	bcData2$"Wrong Way Female"
bcData2$"Helmet" = bcData2$"Helmet Male" +	bcData2$"Helmet Female"

fullSummary <- as.data.frame(matrix(ncol=0, nrow=18))

rownames(fullSummary) <- c("Total Count", "Male Count", "Female Count",  "With Traffic", "With Traffic Male",	"With Traffic Female",	"Sidewalk", "Sidewalk Male", "Sidewalk Female", "Wrong Way", "Wrong Way Male",	"Wrong Way Female", 	"Helmet", "Helmet Male",	"Helmet Female",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses")

fullSummary$Totals <- colSums(bcData2[c("Total Count", "Male Count", "Female Count",  "With Traffic", "With Traffic Male",	"With Traffic Female",	"Sidewalk", "Sidewalk Male", "Sidewalk Female", "Wrong Way", "Wrong Way Male",	"Wrong Way Female", 	"Helmet", "Helmet Male",	"Helmet Female",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses")])

grandTotal <- fullSummary["Total Count", "Totals"]
grandTotal

fullSummary$Percent <- round((fullSummary$Totals / grandTotal)*100, digits = 2)
fullSummary

#Add BPH based on Bikes Per Hour Projected???

```

## Compute subtotals

```{r compute_subtotals}

facilities <- unique(bcData2$"Facility for Street")

for ( f in facilities){ 
  facilitySubset <- bcData2[bcData2$"Facility for Street" == f, ]
  
  facilitySubsetSheet <- createSheet(wb, f)
  
 #rownames(facilitySubset) <- facilitySubset$"Record ID"
  
  addDataFrame(facilitySubset, sheet=facilitySubsetSheet, startColumn=1, row.names=FALSE, 
                colnamesStyle = TABLE_COLNAMES_STYLE, rownamesStyle = TABLE_ROWNAMES_STYLE)

  fullSummary[, f] <- colSums(facilitySubset[c("Total Count", "Male Count", "Female Count",  "With Traffic", "With Traffic Male",	"With Traffic Female",	"Sidewalk", "Sidewalk Male", "Sidewalk Female", "Wrong Way", "Wrong Way Male",	"Wrong Way Female", 	"Helmet", "Helmet Male",	"Helmet Female",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses")])

  subGrandTotal <- fullSummary[1, f]

  fieldName <- paste(f, "Percent", sep=" ")
  
  fullSummary[, fieldName] <- round((fullSummary[, f] / subGrandTotal)*100, digits = 2)

  #Add BPH based on Bikes Per Hour Projected???

  }

#myres <- fullSummary[fullSummary$VarName =="Sidewalk",]
#myres <- fullSummary[fullSummary$VarName =="Helmet",]
#myres <- fullSummary[fullSummary$VarName =="Wrong Way",]

# facilitySubset <- bcData2[bcData2$"Facility for Street" == f, ]
  facilitySubset
```


## Generate short summary

```{r compute_subtotals}

shortSummary <- fullSummary

shortSummary <- shortSummary[c("Total Count", "Male Count", "Female Count",  "With Traffic", "Sidewalk", "Wrong Way", "Helmet",	"Bikes Share Users", 	"Bike Share Sidewalk Wrong Way",	"Bikes on Buses"),]

# Transpose rows into columns to get a dataframe of 4 columns and 573 rows
#shortSummary <- data.frame(t(shortSummary))

# Remove the periods that were added in the column names by default
#names(shortSummary) = gsub(pattern = "\\.", 
#                     replacement = " ", x = names(shortSummary))


shortSummary

```


## Export to Excel - End

```{r export_to_excel_end}

addDataFrame(shortSummary, sheet=shortSummarySheet, startColumn=1, row.names=TRUE, colnamesStyle = TABLE_COLNAMES_STYLE, rownamesStyle = TABLE_ROWNAMES_STYLE)

addDataFrame(fullSummary, sheet=fullSummarySheet, startColumn=1, row.names=TRUE, colnamesStyle = TABLE_COLNAMES_STYLE, rownamesStyle = TABLE_ROWNAMES_STYLE)

bcDataSheet = createSheet(wb, "All Data - Cleaned")
addDataFrame(bcData, sheet=bcDataSheet, startColumn=1, row.names=FALSE, colnamesStyle = TABLE_COLNAMES_STYLE, rownamesStyle = TABLE_ROWNAMES_STYLE)


bcDataOrigSheet = createSheet(wb, "Original Data")
addDataFrame(bcDataOrig, sheet=bcDataOrigSheet, startColumn=1, row.names=FALSE, colnamesStyle = TABLE_COLNAMES_STYLE, rownamesStyle = TABLE_ROWNAMES_STYLE)


saveWorkbook(wb, fileToWrite)

```













